---
title: 'Loan Data from Prosper'
author: 'Alexander Barnes'
output: html_document
---
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(scales)
library(gridExtra)
```

```{r echo=FALSE, Load_the_Data}
loans <- read.csv('prosperLoanData.csv')
```

This data set contains 113,937 loans with 81 variables on each loan, including
loan amount, borrower rate (or interest rate), current loan status, borrower
income, borrower employment status, borrower credit history, and the latest
payment information. The data set was last updated on March 11, 2014.

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
dim(loans)
```

The original dataset contains almost 114,000 loans with 81 variables. In this
analysis, we will limit the amount of variables we analyze to 13 rather
than 81 variables. The variables below were selected out of an interest to see
any relationships and trends between them.

```{r echo=FALSE}
cols <- c('Term', 'LoanStatus', 'BorrowerRate', 'ProsperScore',
          'EmploymentStatus', 'IsBorrowerHomeowner', 'CreditScoreRangeLower',
          'CreditScoreRangeUpper', 'OpenCreditLines', 'AmountDelinquent',
          'DebtToIncomeRatio', 'StatedMonthlyIncome', 'MonthlyLoanPayment')

loans <- loans[,cols]

dim(loans)
```

```{r echo=FALSE}
str(loans)
```

```{r echo=FALSE}
summary(loans)
```


```{r echo=FALSE}
qplot(Term, data = loans, binwidth = 1)
```

```{r echo=FALSE}
table(loans$Term)
```

There are only three distinct values to the loan term (12, 36, and 60 months).
Going forward, we will treat the Term variable as a factor due to its
categorical nature.

```{r echo=FALSE}
loans$Term <- as.factor(loans$Term)
qplot(Term, data = loans)
```

We can clearly see that most loans have a term of 36 months, and it's much less
common for a loan to have a term of 12 months.

```{r echo=FALSE}
qplot(LoanStatus, data = loans) +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))
```

```{r echo=FALSE}
table(loans$LoanStatus)
```

Most of the loans are either Current or Completed, but there are also many that
are Charged off or Defaulted. Of the loans that are Past Due, the highest number
are past due by 1-15 days. There are hardly any loans that are categorized as
\> 120 days past due. My theory is that this is due to the loans being
considered Charged off shortly after they become \> 120 days past due.

```{r echo=FALSE}
qplot(MonthlyLoanPayment, data = loans, binwidth = 50) +
  scale_x_continuous(breaks = seq(0, 2000, 250))
```

```{r echo=FALSE}
summary(loans$MonthlyLoanPayment)
```

Monthly loan payment is skewed to the right, with a minimum of \$0 and a maximum
of \$2251.50. The minimum of \$0 is interesting, as I expected all monthly loan
payments to be greater than \$0. I'm curious which factors contribute to higher
monthly payments.

```{r echo=FALSE}
summary(subset(loans, MonthlyLoanPayment == 0))
```

Subsetting the data to view loans with monthly payments of $0, we can see that
categories such as CreditScoreRangeLower and StatedMonthlyIncome are lower than
the overall median. Interestingly, the ProsperScore is higher than average for
this subset.

```{r echo=FALSE}
summary(subset(loans, MonthlyLoanPayment > 2000))
```

Subsetting the data to view loans with monthly payments over \$2000, it seems
that the borrower generally has a high stated monthly income,  has no amount
delinquent, is employed, and has a high credit score. Also, the terms are all
12 months and the borrower rates are all under 0.15. The borrowers in these
loans appear to be "low risk" borrowers as the higher median Prosper Score
indicates.

```{r echo=FALSE}
loans$ProsperScore <- as.factor(loans$ProsperScore)
qplot(x = ProsperScore, data = subset(loans, !is.na(ProsperScore)))
```

The Prosper Score appears to be a normal distribution. The plot above excludes
any 'NA' values. The Prosper Score is supposed to range from 1-10, so the scores
of 11 were unexpected. I'm hesitant to treat a score of 11 as the best score
since it may be an entry error. Let us subset the data to see if there is a
trend in the loans with a Prosper Score of 11.

```{r echo=FALSE}
summary(subset(loans, ProsperScore == 11))
```

The summary of loans with a Prosper Score of 11 do seem to indicate borrowers
that are low risk. The median StatedMonthlyIncome and OpenCreditLines are higher
than the overall median.

```{r echo=FALSE}
qplot(BorrowerRate, data = loans, binwidth = 0.01)
```

Borrower Rate has a fairly normal distribution, but there is a second peak
around 0.31. The Prosper Score and Borrower Rate are both generated values, and
I'm interested in exploring which variables affect these values.

```{r echo=FALSE}
qplot(EmploymentStatus, data = loans) +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))
```

Almost all of the borrowers are employed. A limitation of this dataset that can
be recognized here is that the data is only for those who applied for a loan and
were approved. The "riskier"" loan applicants (unemployed) that were denied
loans are not a part of this dataset.

I also wonder if leaving the response blank, or the unclear responses of
"Not available" and "Other" affect the borrower's interest rate and Prosper
score.

```{r echo=FALSE}
qplot(IsBorrowerHomeowner, data = loans)
```

```{r echo=FALSE}
qplot(StatedMonthlyIncome, data = loans) +
  scale_x_continuous(limits = c(0, quantile(loans$StatedMonthlyIncome, 0.99)))
```

```{r echo=FALSE}
qplot(StatedMonthlyIncome, data = loans, binwidth = 0.05,
      main = 'Stated Monthly Income (Log10)') +
  scale_x_continuous(trans = log10_trans(),
                     limits = c(1, quantile(loans$StatedMonthlyIncome, 0.99)))
```

```{r echo=FALSE}
qplot(DebtToIncomeRatio,
      data = subset(loans, !is.na(DebtToIncomeRatio)),
      xlim = c(0, quantile(loans$DebtToIncomeRatio, 0.99, na.rm = T)),
      binwidth = .01)
```

There is an even split between homeowners and non-homeowners. Does being a
homeowner affect the Prosper Score?

The stated monthly income is skewed to the right, but the log transformation
produces a normal distribution. There are very large outliers in the stated
monthly income data so the top 1% of data was trimmed for the plot.

Similarly, a majority of the Debt to Income Ratios were between 0 and 1 with
some values at the max of 10.01. The top 1% of data was trimmed for the plot
above, showing a fairly normal distribution with some skew to the right.

```{r echo=FALSE}
summary(subset(loans, DebtToIncomeRatio > 10))
```

Subsetting the data by Debt to Income Ratios greater than 10, we begin to see
riskier Prosper scores, more Charged off or Defaulted loans, and much lower
levels of full-time employment and stated monthly incomes.

```{r echo=FALSE}
lower <- qplot(CreditScoreRangeLower, data = loans, binwidth = 1)

upper <- qplot(CreditScoreRangeUpper, data = loans, binwidth = 1)

grid.arrange(lower, upper)
```

```{r echo=FALSE}
lower <- lower +
  scale_x_continuous(
    limits = c(quantile(loans$CreditScoreRangeLower, .01, na.rm = T), 1000))

upper <- upper +
  scale_x_continuous(
    limits = c(quantile(loans$CreditScoreRangeUpper, .01, na.rm = T), 1000))

grid.arrange(lower, upper)
```

```{r echo=FALSE}
table(loans$CreditScoreRangeLower)
```

```{r echo=FALSE}
table(loans$CreditScoreRangeUpper)
```

The credit score variables both have normal distributions that look identical,
but shifted on the x-axis. Looking at the tables of the values, we can see that
the lower credit scores all end in 0 and the upper scores all end in 9. The
scores also increment by a factor of 20 and the upper credit scores are all
higher than the lower scores by exactly 19. Since these values are clearly
related, we will use the CreditScoreRangeUpper as the credit score variable
going forward.

```{r echo=FALSE}
summary(loans$CreditScoreRangeUpper)
```

According to [ValuePenguin](https://www.valuepenguin.com/average-credit-score),
the average credit score in the U.S. in 2014 was 693. Here we can see that the
mean and median of the CreditScoreRangeUpper in our data set is higher than
this, but again this is the upper value on the credit score range. I'm surprised
to see here that the upper credit scores for those approved for loans is not
much higher than the average in the U.S. in 2014.

```{r echo=FALSE}
qplot(OpenCreditLines, data = loans, binwidth = 1)
```

```{r echo=FALSE}
qplot(AmountDelinquent, data = loans,
      binwidth = 50) +
  scale_x_continuous(limits = c(0,
                                quantile(loans$AmountDelinquent, 0.99,
                                         na.rm = T))) +
  scale_y_continuous(limits = c(0, 1500))
```

```{r}
qplot(AmountDelinquent, data = subset(loans, AmountDelinquent > 0),
      binwidth = 0.1) +
  scale_x_log10(breaks = c(10, 100, 1000, 10000))
```


The number of open credit lines is a fairly normal distribution. Most people
have less than 20 open credit lines. The distribution of amount delinquent is,
alternatively, clearly skewed to the right. Most people are not deliquent at
all. A log transformation of the AmountDelinquent data produces a normal
distribution.

# Univariate Analysis

### What is the structure of your dataset?

The data set being analyzed contains 113,937 loans with 13 variables (Term,
LoanStatus, MonthlyLoanPayment, ProsperScore, BorrowerRate, EmploymentStatus,
IsBorrowerHomeowner, StatedMonthlyIncome, DebtToIncomeRatio,
CreditScoreRangeLower, CreditScoreRangeUpper, OpenCreditLines, and
AmountDelinquent).

The original data set contained 81 variables, but in this project we are
focusing on 13 variables of interest.

Some initial observations from the data are:
- Most loans have a 36 month term
- Borrower rate peaks around 0.15, but also again around 0.31
- The median upper credit score is about the same as the average U.S credit
score
- There are extreme values in the debt to income ratio at 10.01
- Some monthly loan payments are much higher than average
- Some Prosper scores are an 11, outside the scale of 1-10

### What is/are the main feature(s) of interest in your dataset?

The main feature of interest is ProsperScore. I want to explore how the borrower
profile such as EmploymentStatus, CreditScoreRangeUpper, and StatedMonthlyIncome
affect the ProsperScore. Also, I want to see how ProsperScore relates to the
loan details such as BorrowerRate, MonthlyLoanPayment, and LoanStatus.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

As stated above, I think the variables relating to the borrower's profile will
all contribute to the ProsperScore. I think the CreditScoreRangeUpper,
StatedMonthlyIncome, and DebtToIncomeRatio will contribute the most. Further,
I think the ProsperScore will correlate with the BorrowerRate and
MonthlyLoanPayment, and will be predictive of the LoanStatus.

### Did you create any new variables from existing variables in the dataset?

I did not create any new variables, but I did alter existing variables. I made
both ProsperScore and Term factor variables since the numerical values fall
within distinct categories.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

I performed a log transformation on the StatedMonthlyIncome and AmountDelinquent
since the data in both plots was skewed to the right. I wanted to see if the log
transformation would normalize the data. The transformed data did create a more
normal looking distribution for both variables.

# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!