---
title: 'Loan Data from Prosper'
author: 'Alexander Barnes'
output: html_document
---
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(scales)
library(gridExtra)
library(psych)
```

```{r echo=FALSE, Load_the_Data}
loans <- read.csv('prosperLoanData.csv')
```

This data set contains 113,937 loans with 81 variables on each loan, including
loan amount, borrower rate (or interest rate), current loan status, borrower
income, borrower employment status, borrower credit history, and the latest
payment information. The data set was last updated on March 11, 2014.

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
dim(loans)
```

The original dataset contains almost 114,000 loans with 81 variables. In this
analysis, we will limit the amount of variables we analyze to 13 rather
than 81 variables. The variables below were selected out of an interest to see
any relationships and trends between them.

```{r echo=FALSE}
cols <- c('Term', 'LoanStatus', 'BorrowerRate', 'ProsperScore',
          'EmploymentStatus', 'IsBorrowerHomeowner', 'CreditScoreRangeLower',
          'CreditScoreRangeUpper', 'OpenCreditLines', 'AmountDelinquent',
          'DebtToIncomeRatio', 'StatedMonthlyIncome', 'MonthlyLoanPayment')

loans <- loans[,cols]

dim(loans)
```

```{r echo=FALSE}
str(loans)
```

Order levels of loan status and employment status by perceived level of benefit
and risk for the loan company (from worst to best).

```{r echo=FALSE}
LoanStatusLevels <- c('ChargedOff', 'Defaulted', 'Cancelled',
                      'Past Due (>120 days)', 'Past Due (91-120 days)',
                      'Past Due (61-90 days)', 'Past Due (31-60 days)',
                      'Past Due (16-30 days)', 'Past Due (1-15 days)',
                      'FinalPaymentInProgress', 'Current', 'Completed')
loans$LoanStatus <- ordered(loans$LoanStatus, levels = LoanStatusLevels)
levels(loans$LoanStatus)
```

```{r echo=FALSE}
EmploymentLevels <- c('Not employed', '', 'Not available', 'Other', 'Retired',
                      'Self-employed', 'Part-time', 'Employed', 'Full-time')
loans$EmploymentStatus <- ordered(loans$EmploymentStatus,
                                  levels = EmploymentLevels)
levels(loans$EmploymentStatus)[levels(loans$EmploymentStatus) == ''] <- 'No answer'
levels(loans$EmploymentStatus)
```

```{r echo=FALSE}
summary(loans)
```


```{r echo=FALSE}
qplot(Term, data = loans, binwidth = 1)
```

```{r echo=FALSE}
table(loans$Term)
```

There are only three distinct values to the loan term (12, 36, and 60 months).
Going forward, we will treat the Term variable as a factor due to its
categorical nature.

```{r echo=FALSE}
loans$Term <- as.factor(loans$Term)
qplot(Term, data = loans)
```

We can clearly see that most loans have a term of 36 months, and it's much less
common for a loan to have a term of 12 months.

```{r echo=FALSE}
qplot(LoanStatus, data = loans) +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))
```

```{r echo=FALSE}
table(loans$LoanStatus)
```

Most of the loans are either Current or Completed, but there are also many that
are Charged off or Defaulted. Of the loans that are Past Due, the highest number
are past due by 1-15 days. There are hardly any loans that are categorized as
\> 120 days past due. My theory is that this is due to the loans being
considered Charged off shortly after they become \> 120 days past due.

```{r echo=FALSE}
qplot(MonthlyLoanPayment, data = loans, binwidth = 50) +
  scale_x_continuous(breaks = seq(0, 2000, 250))
```

```{r echo=FALSE}
summary(loans$MonthlyLoanPayment)
```

Monthly loan payment is skewed to the right, with a minimum of \$0 and a maximum
of \$2251.50. The minimum of \$0 is interesting, as I expected all monthly loan
payments to be greater than \$0. I'm curious which factors contribute to higher
monthly payments.

```{r echo=FALSE}
summary(subset(loans, MonthlyLoanPayment == 0))
```

Subsetting the data to view loans with monthly payments of $0, we can see that
categories such as CreditScoreRangeLower and StatedMonthlyIncome are lower than
the overall median. Interestingly, the ProsperScore is higher than average for
this subset.

```{r echo=FALSE}
summary(subset(loans, MonthlyLoanPayment > 2000))
```

Subsetting the data to view loans with monthly payments over \$2000, it seems
that the borrower generally has a high stated monthly income,  has no amount
delinquent, is employed, and has a high credit score. Also, the terms are all
12 months and the borrower rates are all under 0.15. The borrowers in these
loans appear to be "low risk" borrowers as the higher median Prosper Score
indicates.

```{r echo=FALSE}
qplot(x = ProsperScore, data = subset(loans, !is.na(ProsperScore)),
      binwidth = 1) +
  scale_x_continuous(breaks = seq(1,11,1))
```

The Prosper Score appears to be a normal distribution. The plot above excludes
any 'NA' values. The Prosper Score is supposed to range from 1-10, so the scores
of 11 were unexpected. I'm hesitant to treat a score of 11 as the best score
since it may be an entry error. Let us subset the data to see if there is a
trend in the loans with a Prosper Score of 11.

```{r echo=FALSE}
summary(subset(loans, ProsperScore == 11))
```

The summary of loans with a Prosper Score of 11 do seem to indicate borrowers
that are low risk. The median StatedMonthlyIncome and OpenCreditLines are higher
than the overall median.

```{r echo=FALSE}
qplot(BorrowerRate, data = loans, binwidth = 0.01)
```

Borrower Rate has a fairly normal distribution, but there is a second peak
around 0.31. The Prosper Score and Borrower Rate are both generated values, and
I'm interested in exploring which variables affect these values.

```{r echo=FALSE}
qplot(EmploymentStatus, data = loans) +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))
```

Almost all of the borrowers are employed. A limitation of this dataset that can
be recognized here is that the data is only for those who applied for a loan and
were approved. The "riskier"" loan applicants (unemployed) that were denied
loans are not a part of this dataset.

I also wonder if leaving the response blank, or the unclear responses of
"Not available" and "Other" affect the borrower's interest rate and Prosper
score.

```{r echo=FALSE}
qplot(IsBorrowerHomeowner, data = loans)
```

```{r echo=FALSE}
qplot(StatedMonthlyIncome, data = loans) +
  scale_x_continuous(limits = c(0, quantile(loans$StatedMonthlyIncome, 0.99)))
```

```{r echo=FALSE}
qplot(StatedMonthlyIncome, data = subset(loans, StatedMonthlyIncome > 0),
      binwidth = 0.05,
      main = 'Stated Monthly Income (Log10)') +
  scale_x_continuous(trans = log10_trans(),
                     limits = c(1, quantile(loans$StatedMonthlyIncome, 0.99)))
```

```{r echo=FALSE}
qplot(DebtToIncomeRatio,
      data = subset(loans, !is.na(DebtToIncomeRatio)),
      xlim = c(0, quantile(loans$DebtToIncomeRatio, 0.99, na.rm = T)),
      binwidth = .01)
```

There is an even split between homeowners and non-homeowners. Does being a
homeowner affect the Prosper Score?

The stated monthly income is skewed to the right, but the log transformation
produces a normal distribution. There are very large outliers in the stated
monthly income data so the top 1% of data was trimmed for the plot.

Similarly, a majority of the Debt to Income Ratios were between 0 and 1 with
some values at the max of 10.01. The top 1% of data was trimmed for the plot
above, showing a fairly normal distribution with some skew to the right.

```{r echo=FALSE}
summary(subset(loans, DebtToIncomeRatio > 10))
```

Subsetting the data by Debt to Income Ratios greater than 10, we begin to see
riskier Prosper scores, more Charged off or Defaulted loans, and much lower
levels of full-time employment and stated monthly incomes.

```{r echo=FALSE}
lower <- qplot(CreditScoreRangeLower, data = loans, binwidth = 1)

upper <- qplot(CreditScoreRangeUpper, data = loans, binwidth = 1)

grid.arrange(lower, upper)
```

```{r echo=FALSE}
lower <- lower +
  scale_x_continuous(
    limits = c(quantile(loans$CreditScoreRangeLower, .01, na.rm = T), 1000))

upper <- upper +
  scale_x_continuous(
    limits = c(quantile(loans$CreditScoreRangeUpper, .01, na.rm = T), 1000))

grid.arrange(lower, upper)
```

```{r echo=FALSE}
table(loans$CreditScoreRangeLower)
```

```{r echo=FALSE}
table(loans$CreditScoreRangeUpper)
```

The credit score variables both have normal distributions that look identical,
but shifted on the x-axis. Looking at the tables of the values, we can see that
the lower credit scores all end in 0 and the upper scores all end in 9. The
scores also increment by a factor of 20 and the upper credit scores are all
higher than the lower scores by exactly 19. Since these values represent a
credit score range that seems to vary by exactly 19 for each borrower, we will
create a new variable CreditScore that is the mean of CreditScoreRangeLower and
CreditScoreRangeUpper.

```{r echo=FALSE}
loans$CreditScore <- (loans$CreditScoreRangeLower+loans$CreditScoreRangeUpper)/2
summary(loans$CreditScore)
```

According to [ValuePenguin](https://www.valuepenguin.com/average-credit-score),
the average credit score in the U.S. in 2014 was 693. Here we can see that the
mean (695) and median (690) of the CreditScore in our data set are right around
this value. I'm surprised to see here that the credit scores for those approved
for loans is not higher than the average in the U.S. in 2014.

We can also see here that credit scores in our data set range from 9.5 to 889.5,
but the actual range for credit scores is 300 to 850. Let's explore some loans
with credit scores outside the normal range.

```{r echo=FALSE}
head(subset(loans, loans$CreditScore < 300), 20)
```

```{r echo=FALSE}
head(subset(loans, loans$CreditScore > 850), 20)
```

These loans do seem to be missing a fair amount of values (there are a lot of
Prosper scores missing), but since the data here do contain values for many of
the variables I will not remove these rows. I will remember to subset the data
for any plots or calculations involving credit score to only include scores from
300 to 850.

```{r echo=FALSE}
loans_subset_CreditScore <- subset(loans, loans$CreditScore >=300 &
                                     loans$CreditScore <= 850)
```

```{r echo=FALSE}
qplot(OpenCreditLines, data = loans, binwidth = 1)
```

```{r echo=FALSE}
qplot(AmountDelinquent, data = loans,
      binwidth = 50) +
  scale_x_continuous(limits = c(0,
                                quantile(loans$AmountDelinquent, 0.99,
                                         na.rm = T))) +
  scale_y_continuous(limits = c(0, 1500))
```

```{r}
qplot(AmountDelinquent, data = subset(loans, AmountDelinquent > 0),
      binwidth = 0.1) +
  scale_x_log10(breaks = c(10, 100, 1000, 10000))
```


The number of open credit lines is a fairly normal distribution. Most people
have less than 20 open credit lines. The distribution of amount delinquent is,
alternatively, clearly skewed to the right. Most people are not deliquent at
all. A log transformation of the AmountDelinquent data produces a normal
distribution.

# Univariate Analysis

### What is the structure of your dataset?

The data set being analyzed contains 113,937 loans with 13 variables (Term,
LoanStatus, MonthlyLoanPayment, ProsperScore, BorrowerRate, EmploymentStatus,
IsBorrowerHomeowner, StatedMonthlyIncome, DebtToIncomeRatio,
CreditScoreRangeLower, CreditScoreRangeUpper, OpenCreditLines, and
AmountDelinquent).

The original data set contained 81 variables, but in this project we are
focusing on 13 variables of interest.

Some initial observations from the data are:
- Most loans have a 36 month term
- Borrower rate peaks around 0.15, but also again around 0.31
- The median upper credit score is about the same as the average U.S credit
score
- There are extreme values in the debt to income ratio at 10.01
- Some monthly loan payments are much higher than average
- Some Prosper scores are an 11, outside the scale of 1-10

### What is/are the main feature(s) of interest in your dataset?

The main feature of interest is ProsperScore. I want to explore how the borrower
profile such as EmploymentStatus, CreditScore, and StatedMonthlyIncome affect
the ProsperScore. Also, I want to see how ProsperScore relates to the loan
details such as BorrowerRate, MonthlyLoanPayment, and LoanStatus.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

As stated above, I think the variables relating to the borrower's profile will
all contribute to the ProsperScore. I think the CreditScore,
StatedMonthlyIncome, and DebtToIncomeRatio will contribute the most. Further,
I think the ProsperScore will correlate with the BorrowerRate and
MonthlyLoanPayment, and will be predictive of the LoanStatus.

### Did you create any new variables from existing variables in the dataset?

I created CreditScore by taking the average of the upper and lower credit
scores, and I made Term a factor variable since the numerical values fell
within three distinct categories. I also ordered the EmploymentStatus and
LoanStatus factors from the perception of best to worst in the eyes of
the loan company.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

I performed a log transformation on the StatedMonthlyIncome and AmountDelinquent
since the data in both plots was skewed to the right. I wanted to see if the log
transformation would normalize the data. The transformed data did create a more
normal looking distribution for both variables.

# Bivariate Plots Section

```{r echo=FALSE, Bivariate_Plots}
no_cor <- c("CreditScoreRangeLower", "CreditScoreRangeUpper", "LoanStatus",
            "EmploymentStatus", "IsBorrowerHomeowner", "Term")
r <- cor(loans_subset_CreditScore[, !(names(loans) %in% no_cor)],
         use = 'complete.obs')
round(r, 2)
```

Borrower rate has a moderate (to strong) negative correlation with Prosper score
and credit score. There are also weak relationships between some of the other
variables. As credit score goes up, Prosper score tends to go up. A higher
monthly loan payment also weakly correlates with a higher stated monthly income
and a lower borrower rate.

```{r echo=FALSE}
no_plot <- c(no_cor, c('AmountDelinquent'))
pairs.panels(loans_subset_CreditScore[,!(names(loans) %in% no_plot)],
             jiggle = T, factor = 1)
```

The Prosper score seems to be moderately correlated with borrower rate
(negatively) and credit score. Credit score also seems to be moderately
correlated with monthly loan payment and negatively with borrower rate.

```{r echo=FALSE}
cat <- c('ProsperScore', 'BorrowerRate', 'LoanStatus', 'EmploymentStatus',
         'IsBorrowerHomeowner')
pairs.panels(loans[,(names(loans) %in% cat)])
```

There don't seem to be any obvious relationships between Prosper score and any
of the categorical variables.

I want to take a closer look at the relationships between Prosper score and
the variables borrower rate, monthly loan payment, credit score, and debt to
income ratio.

```{r echo=FALSE}
ggplot(aes(x = ProsperScore, y = BorrowerRate),
       data = subset(loans, !is.na(ProsperScore))) +
  geom_jitter(alpha = 0.1)
```

As Prosper score increases, borrower rate tends to decrease. At the low end and
high end of the x-axis the variation in borrower rate is lower. In the middle of
the plot (Prosper score of 4-8) we can see a lot more variation in the borrower
rate.

```{r echo=FALSE}
ggplot(aes(x = ProsperScore, y = MonthlyLoanPayment),
       data = subset(loans, !is.na(ProsperScore))) +
  geom_jitter(alpha = 0.2)
```

The slight upward trend in monthly loan payment as Prosper score increases can
be seen here. It's also fairly clear that as Prosper score increases, the
biggest change to monthly loan payments is a wider range and greater volume of
values. We don't see many monthly loan payments greater than $750 for Prosper
scores of 1-3, but the distribution of monthly loan payments for Prosper scores
greater than 3 do not differ much as the score increases. Basically, borrowers
with low Prosper scores do not get loans with high monthly payments.

> Note that for the first two plots I placed ProsperScore on the x-axis. This is
because I believe ProsperScore is used to determine the borrower rate and the 
monthly loan payment. For the next two plots I'll place Prosper score on the
y-axis since I believe credit score and debt to income ratio are variables that
contribute to the calculation of the Prosper score.

```{r echo=FALSE}
ggplot(aes(x = CreditScore, y = ProsperScore),
       data = subset(loans_subset_CreditScore, !is.na(ProsperScore))) +
  geom_jitter()
```

Similarly to the borrower rate, a low Prosper score typically corresponds to a
low credit score and a high Prosper score with a high credit score, but the
middle range of Prosper scores correspond to a wide range of credit scores. In
other words, it's unlikely for a borrower with a low credit score to have a
high Prosper score, and vice versa for a borrower with a high credit score. If
the borrower has a middle range credit score (650-750), their Prosper score
can range vastly.

```{r echo=FALSE}
ggplot(aes(x = DebtToIncomeRatio, y = ProsperScore),
       data = subset(loans, !is.na(ProsperScore))) +
  geom_jitter()
```

```{r echo=FALSE}
ggplot(aes(x = DebtToIncomeRatio, y = ProsperScore),
       data = subset(loans, !is.na(ProsperScore))) +
  geom_jitter(alpha = 0.1) +
  scale_x_continuous(limits = c(0, 2.5))
```

Here we can see the slight negative correlation between debt to income ratio
and Prosper score.

Let us also take a look at borrower rate vs. monthly loan payment and credit
score as these variables had weak to moderate correlations with borrower rate.

```{r echo=FALSE}
qplot(x = MonthlyLoanPayment, y = BorrowerRate, data = loans)
```

```{r echo=FALSE}
ggplot(aes(x = MonthlyLoanPayment, y = BorrowerRate), data = loans) +
  geom_point(alpha = 1/50) +
  scale_x_continuous(limits = c(0, 1250))
```

Changing the plot limits and adding transparency reveals clear sections of the
plot where borrower rate increases linearly as the monthly loan payment
increases. Also, the highest borrower rates tend to correspond to lower monthly
payments.

```{r echo=FALSE}
ggplot(aes(x = CreditScore, y = BorrowerRate),
      data = loans_subset_CreditScore) +
  geom_jitter()
```

The correlation between credit score and borrower rate is clear here - as credit
score increases, borrower rate decreases.

Next, I'll look at how Prosper score and borrower rate may fluctuate based on
the categorical variables (loan status, employment, and homeowners). It is
important to remember here that the loan status and employment status levels
were ordered by my own thoughts on levels of borrower risk and ideal loan
outcomes.

```{r echo=FALSE}
ggplot(aes(x = LoanStatus, y = ProsperScore),
       data = subset(loans, !is.na(loans$LoanStatus))) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))
```

```{r echo=FALSE}
ggplot(aes(x = LoanStatus, y = BorrowerRate),
       data = subset(loans, !is.na(loans$LoanStatus))) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))
```

As mentioned above, it's important to remember that the order of these levels
was determined by my own judgment. There is a slight upward trend in Prosper
score across loan status, but the defaulted loans Prosper score values seem to
throw off this trend. There also may be a slight downward trend in borrower rate
across loan status, but it is not strong at all.

```{r echo=FALSE}
ggplot(aes(x = EmploymentStatus, y = ProsperScore),
       data = loans) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))
```

```{r echo=FALSE}
ggplot(aes(x = EmploymentStatus, y = BorrowerRate),
       data = loans) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))
```

Lower average Prosper scores can be seen for borrowers with employment statuses
of not employed, other, and self-employed, while higher average scores
correspond with retired, part-time and full-time.

The borrower rate is generally higher for those that are not employed, but
otherwise there isn't much difference across employment levels.

```{r echo=FALSE}
ggplot(aes(x = IsBorrowerHomeowner, y = ProsperScore),
       data = loans) +
  geom_boxplot()
```

```{r echo=FALSE}
ggplot(aes(x = IsBorrowerHomeowner, y = BorrowerRate),
       data = loans) +
  geom_boxplot()
```

The borrower's status as a homeowner does not appear to affect the Prosper score
at all, but not owning a home does seem to correspond with a higher borrower
rate.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!