---
title: 'Loan Data from Prosper'
author: 'Alexander Barnes'
output: html_document
---
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(scales)
```

```{r echo=FALSE, Load_the_Data}
loans <- read.csv('prosperLoanData.csv')
```

This data set contains 113,937 loans with 81 variables on each loan, including
loan amount, borrower rate (or interest rate), current loan status, borrower
income, borrower employment status, borrower credit history, and the latest
payment information.

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
dim(loans)
```

The original dataset contains almost 114,000 loans with 81 variables. In this
analysis, we will limit the amount of variables we analyze to 13 rather
than 81 variables. The variables below were selected out of an interest to see
any relationships and trends between them.

```{r echo=FALSE, Univariate_Plots}
cols <- c('Term', 'LoanStatus', 'BorrowerRate', 'ProsperScore',
          'EmploymentStatus', 'IsBorrowerHomeowner', 'CreditScoreRangeLower',
          'CreditScoreRangeUpper', 'OpenCreditLines', 'AmountDelinquent',
          'DebtToIncomeRatio', 'StatedMonthlyIncome', 'MonthlyLoanPayment')

loans <- loans[,cols]

dim(loans)
```

```{r echo=FALSE, Univariate_Plots}
str(loans)
```

```{r echo=FALSE, Univariate_Plots}
summary(loans)
```


```{r echo=FALSE, Univariate_Plots}
qplot(Term, data = loans, binwidth = 1)
```

```{r echo=FALSE, Univariate_Plots}
table(loans$Term)
```

There are only three distinct values to the loan term (12, 36, and 60 months).
Going forward, we will treat the Term variable as a factor due to its
categorical nature.

```{r echo=FALSE, Univariate_Plots}
loans$Term <- as.factor(loans$Term)
qplot(Term, data = loans)
```

We can clearly see that most loans have a term of 36 months, and it's much less
common for a loan to have a term of 12 months.

```{r echo=FALSE, Univariate_Plots}
qplot(LoanStatus, data = loans) +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))
```

```{r echo=FALSE, Univariate_Plots}
table(loans$LoanStatus)
```

Most of the loans are either Current or Completed, but there are also many that
are Charged off or Defaulted. Of the loans that are Past Due, the highest number
are past due by 1-15 days. There are hardly any loans that are categorized as
\> 120 days past due. My theory is that this is due to the loans being
considered Charged off shortly after they become \> 120 days past due.

```{r echo=FALSE, Univariate_Plots}
qplot(MonthlyLoanPayment, data = loans, binwidth = 50) +
  scale_x_continuous(breaks = seq(0, 2000, 250))
```

```{r echo=FALSE, Univariate_Plots}
summary(loans$MonthlyLoanPayment)
```

Monthly loan payment is skewed to the right, with a minimum of \$0 and a maximum
of \$2251.50. The minimum of \$0 is interesting, as I expected all monthly loan
payments to be greater than \$0. I'm curious which factors contribute to higher
monthly payments.

```{r echo=FALSE, Univariate_Plots}
loans$ProsperScore <- as.factor(loans$ProsperScore)
qplot(x = ProsperScore, data = subset(loans, !is.na(ProsperScore)))
```

The Prosper Score appears to be a normal distribution. The plot above excludes
any 'NA' values. The Prosper Score is supposed to range from 1-10, so the scores
of 11 were unexpected.

```{r echo=FALSE, Univariate_Plots}
qplot(BorrowerRate, data = loans, binwidth = 0.01)
```

Borrower Rate has a fairly normal distribution, but there is a second peak
around 0.31. The Prosper Score and Borrower Rate are both generated values, and
I'm interested in exploring which variables affect these values.

```{r echo=FALSE, Univariate_Plots}
qplot(EmploymentStatus, data = loans) +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))
```

Almost all of the borrowers are employed. A limitation of this dataset that can
be recognized here is that the data is only for those who applied for a loan and
were approved. The "riskier"" loan applicants (unemployed) that were denied
loans are not a part of this dataset.

I also wonder if leaving the response blank, or the unclear responses of
"Not available" and "Other" affect the borrower's interest rate and Prosper
score.

```{r echo=FALSE, Univariate_Plots}
qplot(IsBorrowerHomeowner, data = loans)
```

```{r echo=FALSE, Univariate_Plots}
qplot(StatedMonthlyIncome, data = loans) +
  scale_x_continuous(limits = c(0, quantile(loans$StatedMonthlyIncome, 0.99)))
```

```{r echo=FALSE, Univariate_Plots}
qplot(StatedMonthlyIncome, data = loans, binwidth = 0.05,
      main = 'Stated Monthly Income (Log10)') +
  scale_x_continuous(trans = log10_trans(),
                     limits = c(1, quantile(loans$StatedMonthlyIncome, 0.99)))
```

```{r echo=FALSE, Univariate_Plots}
qplot(DebtToIncomeRatio,
      data = subset(loans, !is.na(DebtToIncomeRatio)),
      xlim = c(0, quantile(loans$DebtToIncomeRatio, 0.99, na.rm = T)),
      binwidth = .01)
```

```{r echo=FALSE, Univariate_Plots}

```

```{r echo=FALSE, Univariate_Plots}

```

```{r echo=FALSE, Univariate_Plots}

```

```{r echo=FALSE, Univariate_Plots}

```

# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?

### What is/are the main feature(s) of interest in your dataset?

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?


# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!